Higher-Order unification for free!

### 2024-04-16 paper/main.tex
help by yves

\newcommand{\Fo}{\texorpdfstring{\ensuremath{\mathcal{F}_{\!o}\xspace}}{Fo}} % space non va
\newcommand{\Ho}{\texorpdfstring{\ensuremath{\mathcal{H}_o}\xspace}{Ho}}

@@

\section{Introduction}
\label{sec:intro}

Meta languages such as Elf~\cite{elf}, Twelf~\cite{twelf},
$\lambda$Prolog~\cite{miller_nadathur_2012}, and
Isabelle~\cite{10.1007/978-3-540-71067-7_7}
have been utilized to specify various
logics~\cite{felty88cade,books/sp/NipkowPW02,10.1007/BF00881873,felty93lics}.
The use of these meta languages facilitates this task in two
key ways. The first and most well-know one is that variable binding and
substitution come for free. %, exploited in all works mentioned above.
The second one is that these meta languages come equipped with some form
of unification, a cornerstone of proof construction and proof search.

% exploited only
%in the notable case of of Higher Order Logic~\cite{books/sp/NipkowPW02}:
%the meta language Isabelle is such a good match for HOL that it could used to
%implement an interactive proof system for the object logic, in addition to
%it specification.\\
The object logic we are interested in is Coq's~\cite{Coq-refman}
Calculus of Inductive Constructions (CIC), and we aim to implement a
form of proof search known as type-class~\cite{wadler89,sozeau08} resolution.
Type-class solvers are unification based on proof search procedures
reminiscent of Prolog, which back-chain lemmas taken
from a database of ``type-class instances''. Given this
analogy with Logic Programming we want to leverage the
Elpi~\cite{tassi:hal-01637063} meta-programming language,
a dialect of $\lambda$Prolog, already used to extend
Coq in various ways~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
In this paper, we focus on one aspect of
this work, precisely \emph{how to reuse the higher-order unification procedure
of the meta language in order to simulate a higher-order logic program
for the object language}.

We take as an example the \coqIn{Decision} and \coqIn{Finite} type classes
from the Stdpp~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
library. The class  \coqIn{Decision}
identifies predicates equipped with a decision procedure, while
\coqIn{Finite} identifies types whose inhabitants can be enumerated in a (finite) list.
The following three type-class instances state that:
1) the type of natural numbers smaller than \coqIn{n}, called \coqIn{fin n},
is finite;
2) the predicate \coqIn{nfact n nf}, relating a natural number
\coqIn{n} to the number of its prime factors \coqIn{nf}, is decidable;
3) the universal closure of a predicate has a decision procedure if
its domain is finite and if the predicate is decidable.

@@

\noindent
The proof found by the solver back-chains on rule 3 (the only rule
about the $\forall$ quantifier), and then solves the premises with
rules 1 and 2 respectively.
Note that rule 3 features a second-order parameter \coqIn{P} that represents
a function of type \coqIn{A ~$\to$~ Prop} (a predicate over \coqIn{A}).
The solver has to infer a value for \coqIn{P} by unifying the conclusion
of rule 3 with the goal, and in particular, it has to solve the unification
problem \coqIn{P x = nfact x 3}. This higher-order problem falls in the so-called 
pattern-fragment \llambda~\cite{miller92jsc} and admits a unique
solution $\customlabel{solution:intro}{\rho}$ that assigns
the term \elpiIn{~$\lambda$~x.nfact x 3} to \coqIn{P}.

In order to implement such a search in Elpi, we shall describe the encoding
of CIC terms and then the encoding of instances as rules.
Elpi comes equipped with
an Higher Order Abstract Syntax (HOAS~\cite{10.1145/53990.54010}) datatype of CIC
terms, called \elpiIn{tm}, that includes (among others) the following
constructors:

@@

\noindent
We now illustrate a naive encoding of the three instances above as higher-order
logic programming rules: capital letters denote rule
parameters; \elpiIn{:-} separates the rule's head from the premises, and
\elpiIn{pi w\ p} introduces a fresh nominal constant \elpiIn{w}
for the premise \elpiIn{p}.

@@

\noindent
In this paper we study a more sophisticated encoding of CIC terms and rules
that, on a first approximation, would reshape \ref{clause:r3} as follows:

@@

\noindent
The root cause of the problems we outlined in this example
is a subtle mismatch between the equational theories of the meta language
and the object language, which in turn makes the 
unification procedures of the meta language weak.
The equational theory of the meta language Elpi encompasses
$\eta\beta$-equivalence and its unification procedure can solve higher-order
problems in the pattern fragment. Although 
the equational theory of CIC is much richer, for efficiency and predictability
reasons, automatic proof search procedures typically employ a unification
procedure that only captures a $\eta\beta$-equivalence and only operates
in \llambda. The similarity is striking, but one needs to exercise some caution in order to
simulate a logic program in CIC using the unification of Elpi.

\paragraph{Contributions}
In this paper we identify a minimal language \Fo{} in which the problems
sketched in the introduction can be formally described.
We detail an encoding of a logic program in \Fo{} to a strongly related
logic program in \Ho (the language of the meta-language) and we show that
the higher-order unification procedure of the meta language \Ue{} can be
efficiently used to simulate a higher-order unification procedure \Uo for
the object language that features $\eta\beta$-conversion. We show how \Uo
can be extended with heuristics to deal with problems outside the pattern
fragment.\\
Section~\ref{sec:problem-statement} formally states the problem and gives the
intuition behind our solution; \cref{sec:simulation} sets up a basic
simulation of first-order logic programs, \cref{sec:llam} and \cref{sec:eta}
extend it to higher-order logic programs in the pattern fragment
while \cref{sec:beta} goes beyond the pattern fragment.
Section \ref{sec:implementation} discusses the implementation in Elpi.\\
The $\lambda$Prolog code discussed in the paper can be accessed at the
address \githubUrl.

@@

Even if we encountered the problem working on CIC, we devise
a minimal setting to ease its study. In this setting, we have
a \Fo{} language (for first order) with a rich equational
theory and a \Ho{} meta language with a simpler one.

@@

\noindent
In order to simulate a \Fo{} logic program in \Ho{}, we compile
each \Fo{} term $s$ in \foUnifPb to a \Ho{} term $t$.
We write this translation as $\C{s}{t}{m}{l}$. The implementation of the compiler
is detailed in \cref{sec:simulation,sec:eta,sec:beta}, here we just point
out that it additionally produces a variable mapping $m$ and a list of links $l$.
The variable map connects unification variables in \Ho to variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$. Links are an accessory piece of information whose
description is deferred to \cref{sec:nutshell}.
We write $\foUnifPb_p = \{~ \foUnifPb_{p_l}, \foUnifPb_{p_r} ~\}$
and $s \in \foUnifPb \Leftrightarrow \exists p, s \in \foUnifPb_p$.

@@

The final step is to decompile each assignment.
Decompiling a term is straightforward since \mapStore is a bijection.
The only complex part concerns the \elpiIn{abs} node. In our simple setting,
the \elpiIn{flam} node carries no additional information (other than the function body),
so each \elpiIn{abs} node can be trivially converted to a \elpiIn{flam} one.
However, in the case of CIC, where lambdas carry the type of the bound variable, one
must store this information somewhere. Note that this information is similar
to the arity of variables; that in CIC, unification variables have a
(function) type, and this type can be used to annotate the lambdas
needed to express their assignment.

@@

In \cref{sec:grounwork}, we introduced the definition of memory map \mapStore.
This map allows to decompile the \Ho substitution to the object language.
Given the higher order nature of CIC the same unification
variable $X$ is used multiple times with different arities.
%
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}

@@

\noindent
However, this encoding has two big limitations. First it is
not always feasible to adopt it for CIC due to the fact that the type system
of the ML is too limited to accommodate for that one the object language,
e.g. CIC can typecheck variadic functions~\cite{cpdt}. \\
Second, the CIC encoding provided by Elpi is primarily utilized for meta
programming, in order to extend the Coq system. Consequently, it must be able to
manipulate terms that are not known in advance
without relying on introspection primitives such as Prolog's
\texttt{functor} and \texttt{arg}. In this context, constants need to live in
an open world, akin to the \elpiIn{string} data type used in the preceding
examples, they cannot be symbols of the ML.

@@

\emph{The approach presented in this paper}
provides a third option that addresses all the concerns
mentioned earlier. It capitalizes on the benefit of not requiring to fully
implement
the unification algorithm of the object language. Instead, it employs the
unification capabilities of the meta language, facilitated by the various links
%we establish
to manage ``problematic'' subterms. As a result of this choice our encoding takes
advantage of indexing data structures and mode analysis for clause filtering.
It it worth mentioning that we only replace terms with variables only
when it is strictly needed, leaving the rest of the term structure intact
and hence indexable by the meta language logic programming engine.
Moreover, the unification process we obtain is ready to take
advantage of potential improvements to the programming engine,
such as tabled search, and apply forms of static analysis for
the meta language, such as determinacy, to the object language.
\todo{mention preliminar experiment show that our prolog is 2x faster than the Coq one} 
Finally, our approach is flexible enough to accommodate different strategies
and heuristics to handle terms outside the pattern fragment, and
it is not tightly coupled with CIC.

>>>>

%\newcommand{\Fo}{\texorpdfstring{\ensuremath{\mathcal{F}_{\!o}\xspace}}{Fo}} % space non va
%\newcommand{\Ho}{\texorpdfstring{\ensuremath{\mathcal{H}_o}\xspace}{Ho}}
\newcommand{\Fo}{\texorpdfstring{\ensuremath{\mathcal{F}\xspace}}{F}} % space non va
\newcommand{\Ho}{\texorpdfstring{\ensuremath{\mathcal{H}}\xspace}{H}}

@@

\section{Introduction}
\label{sec:intro}
We aim to implement a
form of proof search known as type-class resolution~\cite{wadler89,sozeau08}
for the Coq's system~\cite{Coq-refman}.
Type-class solvers are unification based proof search procedures
reminiscent of Prolog, which back-chain lemmas taken
from a database of ``type-class instances''. Given this
analogy with Logic Programming we want to leverage the
Elpi meta-programming language~\cite{tassi:hal-01637063},
a dialect of $\lambda$Prolog, already used to extend
Coq in various ways~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.

The use of meta languages facilitates the implementation of
proof systems in two key ways. The first and most well-know one is that variable binding and
substitution come for free. %, exploited in all works mentioned above.
The second one is that these meta languages come equipped with some form
of unification, a cornerstone of proof construction and proof search.
% exploited only
%in the notable case of of Higher Order Logic~\cite{books/sp/NipkowPW02}:
%the meta language Isabelle is such a good match for HOL that it could used to
%implement an interactive proof system for the object logic, in addition to
%it specification.\\
In this paper, we focus on one aspect of
this work, precisely \emph{how to reuse the higher-order unification procedure
of the meta language in order to simulate a higher-order logic program
for the object language}.

We take as an example the \coqIn{Decision} and \coqIn{Finite} type classes
from the Stdpp
library~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}.
The class  \coqIn{Decision}
identifies predicates equipped with a decision procedure, while
\coqIn{Finite} identifies types whose inhabitants can be enumerated in a (finite) list.
The following three type-class instances state that:
1) the type of natural numbers smaller than \coqIn{n}, called \coqIn{fin n},
is finite;
2) the predicate \coqIn{nfact n nf}, relating a natural number
\coqIn{n} to the number of its prime factors \coqIn{nf}, is decidable;
3) the universal closure of a predicate has a decision procedure if
its domain is finite and if the predicate is decidable.

@@

\noindent
The proof found by the solver back-chains on rule 3 (the only rule
about the $\forall$ quantifier), and then solves the premises with
rules 1 and 2 respectively.
Note that rule 3 features a second-order parameter \coqIn{P} that represents
a function of type \coqIn{A ~$\to$~ Prop} (a 1-argument predicate over \coqIn{A}).
The solver has to infer a value for \coqIn{P} by unifying the conclusion
of rule 3 with the goal, and in particular, it has to solve the unification
problem \coqIn{P x = nfact x 3}. This higher-order problem falls in the so-called 
pattern-fragment and admits a unique
solution $\customlabel{solution:intro}{\rho}$ that assigns
the term \elpiIn{~$\lambda$~x.nfact x 3} to \coqIn{P}.
In the rest of this paper shall refer to the pattern-fragment as
\llambda~\cite{miller92jsc}.

In order to implement such a search in Elpi, we shall describe the encoding
of Coq terms and then the encoding of instances as rules.
Elpi comes equipped with
an Higher Order Abstract Syntax (HOAS~\cite{10.1145/53990.54010}) datatype of
terms, called \elpiIn{tm}, that includes (among others) the following
constructors:

@@

\noindent
We now illustrate a naive encoding of the three instances above as higher-order
logic programming rules: capital letters denote rule
parameters; \elpiIn{:-} separates the rule's head from the premises, and
<<\elpiIn{pi w\ p}>> introduces a fresh nominal constant \elpiIn{w}
for the premise \elpiIn{p}.

@@

\noindent
The problem above has no solution in a first-order language
where \elpiIn{P} can only stand for a predicate symbol, while in a higher-order
language \elpiIn{P} can stand for any 1-argument predicate.

In this paper we study a more sophisticated encoding of Coq terms and rules
that, on a first approximation, would reshape \ref{clause:r3} as follows:

@@

\noindent
The root cause of the problems we outlined in this example
is a subtle mismatch between the equational theories of the meta language
and the object language, which in turn makes the 
unification procedures of the meta language weak.
The equational theory of the meta language Elpi encompasses
$\eta\beta$-equivalence and its unification procedure can solve higher-order
problems in the pattern fragment. Although 
the equational theory of Coq is much richer, for efficiency and predictability
reasons, automatic proof search procedures typically employ a unification
procedure that only captures a $\eta\beta$-equivalence and only operates
in \llambda. The similarity is striking, but one needs to exercise some caution in order to
simulate a higher-order logic program in Coq using the unification of Elpi.

\paragraph{Contributions}
In this paper we identify a minimal language \Fo{} in which the problems
sketched in the introduction can be formally described.
We detail an encoding of a logic program in \Fo{} to a strongly related
logic program in \Ho (the language of the meta-language) and we show that
the higher-order unification procedure of the meta language \Ue{} can be
efficiently used to simulate a higher-order unification procedure \Uo for
the object language that features $\eta\beta$-conversion. We show how \Uo
can be extended with heuristics to deal with problems outside the pattern
fragment.\\
Section~\ref{sec:problem-statement} states the problem formally and gives the
intuition behind our solution; \cref{sec:simulation} sets up a basic
simulation of first-order logic programs, \cref{sec:llam} and \cref{sec:eta}
extend it to higher-order logic programs in the pattern fragment
while \cref{sec:beta} goes beyond the pattern fragment.
Section \ref{sec:implementation} discusses the implementation in Elpi.\\
The $\lambda$Prolog code discussed in the paper can be accessed at the
address \githubUrl.

@@

Even if we encountered the problem working on Coq's Calculus of Inductive
Constructions, we devise
a minimal setting to ease its study. In this setting, we have
a \Fo{} language (for first order) with a rich equational
theory and a \Ho{} meta language with a simpler one.

@@

\noindent
In order to simulate a \Fo{} logic program in \Ho{}, we compile
each \Fo{} term $s$ in \foUnifPb to a \Ho{} term $t$.
We write this translation as $\C{s}{t}{m}{l}$. The implementation of the compiler
is detailed in \cref{sec:simulation,sec:eta,sec:beta}, here we just point
out that it additionally produces a variable map $m$ and a list of links $l$.
The variable map connects unification variables in \Ho to variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$. Links are an accessory piece of information whose
description is deferred to \cref{sec:nutshell}.
We write $\foUnifPb_p = \{~ \foUnifPb_{p_l}, \foUnifPb_{p_r} ~\}$
and $s \in \foUnifPb \Leftrightarrow \exists p, s \in \foUnifPb_p$.

@@

The final step is to decompile each assignment.
Decompiling a term is straightforward since \mapStore is a bijection.
The only complex part concerns the \elpiIn{abs} node. In our simple setting,
the \elpiIn{flam} node carries no additional information (other than the function body),
so each \elpiIn{abs} node can be trivially converted to a \elpiIn{flam} one.
However, in the case of Coq, where lambdas carry the type of the bound variable, one
must store this information somewhere. Note that this information is similar
to the arity of variables; that in Coq, unification variables have a
(function) type, and this type can be used to annotate the lambdas
needed to express their assignment.

@@

In \cref{sec:grounwork}, we introduced the definition of memory map \mapStore.
This map allows to decompile the \Ho substitution to the object language.
Given the higher order nature of Coq the same unification
variable $X$ is used multiple times with different arities.
%
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}

@@

\noindent
However, this encoding has two big limitations. First it is
not always feasible to adopt it for Coq due to the fact that the type system
of the ML is too limited to accommodate for that one the object language,
e.g. Coq can typecheck variadic functions~\cite{cpdt}. \\
Second, the encoding of Coq terms provided by Elpi is primarily utilized for meta
programming, in order to extend the Coq system. Consequently, it must be able to
manipulate terms that are not known in advance
without relying on introspection primitives such as Prolog's
\texttt{functor} and \texttt{arg}. In this context, constants need to live in
an open world, akin to the \elpiIn{string} data type used in the preceding
examples, they cannot be symbols of the ML.

@@

\emph{The approach presented in this paper}
provides a third option that addresses all the concerns
mentioned earlier. It capitalizes on the benefit of not requiring to fully
implement
the unification algorithm of the object language. Instead, it employs the
unification capabilities of the meta language, facilitated by the various links
%we establish
to manage ``problematic'' subterms. As a result of this choice our encoding takes
advantage of indexing data structures and mode analysis for clause filtering.
It it worth mentioning that we only replace terms with variables only
when it is strictly needed, leaving the rest of the term structure intact
and hence indexable by the meta language logic programming engine.
Moreover, the unification process we obtain is ready to take
advantage of potential improvements to the programming engine,
such as tabled search, and apply forms of static analysis for
the meta language, such as determinacy, to the object language.
\todo{mention preliminar experiment show that our prolog is 2x faster than the Coq one} 
Finally, our approach is flexible enough to accommodate different strategies
and heuristics to handle terms outside the pattern fragment, and
it is not tightly coupled with Coq.
